### 2024-02-20 latex/HP1_comm.tex
new files

\input{PreambleComm}
\begin{document}
\pagestyle{HPed}
\begin{ekdosis}
\SetLineation{lineation = none,}

@@

%%%%%%%%%%
\subsection*{1.1}
\begin{translation}[hp01_001]
Homage to the glorious Ādinātha who taught the science of Haṭhayoga which is like a splendid stairway for one who wants to climb to the lofty royal terrace.%??Add note re meaning of "science".JM: would much prefer "science" to "system" (vidyā doesn't mean system). I changed "ladder" to "stairway" (a bit more poetic). I'd also prefer "which is like a splendid stairway for one who wants to climb to the lofty royal terrace."
%JB: I much prefer Jim's alternative translation (i.e., "which is like...). But I think we should avoid translating vidyā as `science' because it implies that there was some antecedent to the scientific method in premodern yoga traditions. Also, in light of the current politics around yoga in India, "science of Haṭha" fosters the rather absurd view that ancient Indian knowledge systems anticipated scientific discoveries of the 20th century.  JM: But "system" is wrong.

@@

\begin{philcomm}[hp01_001]
%?? JM: now saudha is preferred for stemmatic reasons too. I suggest the following shortened form of the note: 
% JB: I prefer the longer note for the digital edition and the shorter one (proposed by Jim) for the printed edition. How many readers will have access to Hanneder 2020 and read the whole article? The longer note nicely summarises a longer discussion in the article.
The \textalpha\ witnesses read \emph{virājate pronnatarājayogam} while several others have either \emph{vibhrājate} or \emph{°rājasaudham} or both. As explained in Hanneder 2020, \emph{virājate} is stylistically preferable and the comparison between the science of \emph{haṭhayoga} and a stairway works only with \emph{°rājasaudham}.

[Place holder] This reading appears to have resulted from an attempt to replace the word \emph{vidyā} with \emph{yoga} in the opening verses of the text.

%The reading \emph{rājasaudha} is preferable for poetical reasons, as explained in Hanneder 2020, p. 128–130. Also, the \emph{Jyotsnā} explains that the \emph{alaṃkāra} here is a comparison (\emph{upamā}), which consists of four elements: (1) a particle expressing the comparison ?? JM: what's happened here? ALso, do we need to repeat what Jürgen has already written elsewhere?
%The Sanskrit poeticians explain that when some elements of a comparison remains unexpressed we get an incomplete (\emph{lupta}) comparison. Often words like \emph{iva} or \emph{yathā} are missing or the common property, but if we read \emph{rājayoga} we lose the \emph{upamāna} (i.e., \emph{rājasaudha}). One part of the comparison should not be missing. But from the perspective of \emph{alaṃkāraśāstra} the verse has a problem that has surely caused the dilemma: it either leaves the \emph{upameya} or the \emph{upamāna} incomplete. Brahmānanda has mentioned both ?? \emph{yathā adhirohiṇī saudhaprāpikā bhavati evaṃ haṭhadīpikā rājayogaprāpikā bhavati}, but had to supply \emph{rājasaudha} because his text reads \emph{rājayoga}. But to have the \emph{upameya} in the text is odd. Even Brahmānanda could only know of the image (\emph{rājasaudha}) from the alternative reading he did not accept. However, it seems likely that the author would have included the \emph{upamāna} in the text, as in our critical text, so that the reader would know that the upper terrace of the palace is an image for \emph{rājayoga}. It is further likely that the substitution of \emph{rājayoga} for \emph{rājasaudha} was the result of the tendency to insert the word \emph{yoga} in the opening verses of the text wherever possible, even where it does not fit, as can be seen in 1.2d (\emph{haṭhayogopadiśyate}) and 1.3b (\emph{rājayogam ajānatām}). In the latter case the poetical image has also been lost.

?? add comment about yoga replacing mārga and saudha here and in the next two verses.

@@

\begin{philcomm}[hp01_003] 
Most witnesses have \emph{rājayogam ajānatām} (`for those ignorant of Rājayoga') in 1.3b. While this reading is well attested by the manuscript transmission of the \emph{Haṭhapradīpikā} and some manuscripts of the \emph{Haṭharatnāvalī} (note the readings of P,T,T1 of the critical edition [Gharote 2009: 5 n. 2] in the testimonia), the reading \emph{rājamārgam ajānatām}, “for those unable to find the royal highway”, is more appropriate to the metaphor of being lost in darkness.

In 1.3d \emph{kṛpākaraḥ} is attested by the most important groups, including \textalpha, of the available manuscripts. The readings \emph{prakāśyate} and \emph{kṣamākaraḥ} are attested by some manuscripts in lower branches of the stemma. Since the context is the author helping yogis who have strayed from the royal path, \emph{kṛpākaraḥ} makes good sense. As Brahmānanda notes, this compound can be understood as one who is compassionate (\emph{kṛpā} + \emph{kara}) or one who is a mine (i.e., a rich source) of compassion (\emph{kṛpā} + \emph{ākara}). In the Devanagari transmission, the \emph{kṣa} of \emph{kṣamākaraḥ} probably arose as a mistake for \emph{kṛ}.       
\end{philcomm}

%%%%%%%%%%
\subsection*{1.4}
\begin{translation}[hp01_004]
In fact, Matsyendra, Gorakṣa, and other [perfected yogis] discovered the science of Haṭha, and the yogi Svātmārāma knows it through their favour.
\end{translation}

@@

\begin{philcomm}[hp01_008]  
The name Allamaprabhudeva (sometimes Allama Prabhu Deva or Allama Prabhudeva in secondary literature) is largely transmitted as \emph{allamaḥ prabhu\-devaḥ}, as though it were two names, although some manuscripts of the \emph{Haṭha\-pradīpikā} have \emph{allamaprabhudevaś ca} (i.e., \getsiglum{V3,V8,V13,V16,V22,N24,N26,Jyo}) and so do some of the \emph{Haṭharatnāvalī} (i.e., P,T,t1 in Gharote 2009: 35 n. 8).

@@

Many witnesses have \emph{khaṇḍa} and \emph{kāpālika} as separate names. However, Khaṇḍa\-kāpālika is well attested. Examples include \emph{Kathāsaritsāgara} 121.5 ff. (check??), \emph{Bṛhatkathāmañjarī} 10.45 (check??) and Vajrapāṇi’s \emph{Laghutantraṭīkā}, p.45 (\emph{vīrāḥ khaṇḍakāpālikādayaś caturviṃśatiḥ}). It may be a derogatory name for a Kāpālika, coined perhaps by an outsider and connoting something like a defective Kāpālika in the sense of a ‘part-time’ Kāpālika. Alternatively, it could simply refer to one who uses a broken skull as a bowl.
\end{philcomm}

@@

%The reading of \emph{samāśrayo} in \etaOne\ is metrically faulty.

Both \emph{°jagatāṃ} and \emph{°yuktānāṃ} are well attested by the collated manuscripts, and the \textalpha\ group omits the second half of this verse.
We have adopted \emph{°jagatāṃ}  because it makes good sense with  \emph{ādhārakamaṭhah} in light of the cosmological notion that the tortoise supports all the worlds. This reading may not have been understood by some and was changed in the vulgate and other witnesses to \emph{°yuktānāṃ} instead.      
\end{philcomm}

@@

%%%%%%%%%%
\subsection*{1.13}
\begin{translation}[hp01_013]
It has a small door and is without cracks, holes and bumps. It is neither too high nor too low in extent and is thickly smeared with cow dung in the proper way. It is clean, free from all annoyances, pleasing on the outside with a verandah, altar and well, surrounded by a wall: these are the characteristics of the yoga hut as taught by the adept practitioners of Haṭha.
\end{translation}

@@

\begin{philcomm}[hp01_013]  
The syntax of this verse is rather problematic. One would expect the features of the hut, which are listed in the first three quarters of the verse, to be in the nominative case. Then, the words \emph{idaṃ lakṣaṇaṃ} in the fourth quarter would refer back to them. However, the compounds in the first three verse-quarters appear to qualify \emph{lakṣaṇa} as though they were adjectives, and this seems to have been the way the verse was composed.

@@

%??(MD: Stemmatically, \emph{piṭaka} ``a basket; a boil" is well attested. Is it perhaps possible to interpret the word as ``a bump"? \emph{a-randhra-garta-piṭaka} would then mean that the floor or walls of the hut are not uneven.) JM: yes, sounds good to me.
%JB Okay I've written the note to reflect this.

@@

%%%%%%%%%%
\subsection*{1.14}
\begin{translation}[hp01_014]
Staying in such a hut, free from all worry, in the way taught by his guru [the yogi] should practise only yoga.
\end{translation}

@@

%%%%%%%%%%
\subsection*{1.19}
\begin{translation}[hp01_019]
Placing the soles of both feet well between the knees and thighs [and] sitting up with the body straight: they call that \emph{svastikāsana}.%JM: correctly is tricky. It could mean thoroughly, completely — I've looked at some other usages of samyak and that understanding seems to work well quite often, as it does here. E.g. VS 3.37ab hastābhyāṃ bandhayet samyak karṇādikaraṇāni ca, VM 58cd rasānāṃ śoṣanaṃ samyak mahāmudrābhidhīyate. Or Brahmānanda ad 1.35 susthiram samyak sthiraṃ
% JB thoroughly doesn't make sense here. And is it possible to 'completely' place the feet between the knees and thighs?
\end{translation}

@@

%%%%%%%%%%
\subsection*{1.20}
\begin{translation}[hp01_020]
[The yogi] should place his right heel on the left side of the [lower] back, and the left [heel] on the right [side], in the same way. This is \emph{gomukhāsana}, which [looks] like a cow's face.
\end{translation}

@@

%%%%%%%%%%
\subsection*{1.21}
\begin{translation}[hp01_021]
 Fixing one foot on one thigh and placing the [other] thigh on the other foot is called \emph{vīrāsana}.
\end{translation}

@@

%%%%%%%%%%
\subsection*{1.22}
\begin{translation}[hp01_022]
Knowers of yoga know that \emph{kūrmāsana} arises by carefully blocking the anus with the ankles crossed.%??Tweak note to reflect change to crossed. 
\end{translation}

@@

%%%%%%%%%%
\subsection*{1.23}
\begin{translation}[hp01_023]
[The yogi] correctly assumes \emph{padmāsana}, inserts the hands between the knees and thighs, places [the hands] on the ground, and remains in the air. This is \emph{kukkuṭāsana}.
\end{translation}

@@

%%%%%%%%%%
\subsection*{1.24}
\begin{translation}[hp01_024]
While maintaining \emph{kukkuṭāsana}, [the yogi] binds the neck with the hands and lies like a tortoise on his back. This is \emph{uttānakūrmāsana}.
\end{translation}

@@

\begin{philcomm}[hp01_024]
The oldest dated manuscript, \etaOne, has \emph{kukkuṭāsanavat kṛtvā}, which is a simpler alternative to the widely attested reading \emph{kukkuṭāsanabandhasthaḥ}, which we have accepted. Since there is no known source for this verse other than the \emph{Haṭhapradīpikā}, we can only assume that the reading of \etaOne\ was an isolated attempt to simplify the syntax.%??JM: I guess we discussed this before, but are we not going to translate āsana names?
%JB if we do, we should put the Sanskrit in round brackets as many enthusiasts use the Sanskrit names.
\end{philcomm}

%%%%%%%%%%
\subsection*{1.25}
\begin{translation}[hp01_025]
Clasping the big toes with the hands and performing the action of drawing a bow as far as the ear is called \emph{dhanurāsana}.
\end{translation}

@@

%%%%%%%%%%
\subsection*{1.26}
\begin{translation}[hp01_026]
[The yogi] should hold the right foot, which is placed at the base of the left thigh, with the [hand of] the right arm, which is wrapped around the outside of the knee, and remain [like that] with his body twisted. This posture was taught by Matsyendranātha.
\end{translation}

@@

%%%%%%%%%%
\subsection*{1.28}
\begin{translation}[hp01_028]
[The yogi] should stretch out the legs on the ground [as straight] as sticks, hold the toes of both feet with the hands, place the forehead upon the knees and remain thus. They call this the back-stretch (\emph{paścimatānam}).%?? āsana name translated here.
\end{translation}

@@

%%%%%%%%%%
\subsection*{1.29}
\begin{translation}[hp01_029]
This back-stretch is the foremost among \emph{āsana}s. It makes the breath flow to the rear (i.e.~in the central channel), increases the digestive fire, makes the belly thin and prevents diseases in men.
\end{translation}

@@

\begin{philcomm}[hp01_029]
 The use of the word \emph{paścima} to mean the central channel is found at \emph{Yogabīja} 95 (\emph{paścimamārgataḥ}), 108 (\emph{paścime pathi}), 117 (\emph{paścimadvāramārgeṇa}) and 121 (\emph{paścimaṃ}). Cf.~the usages of \emph{paścima\-mārga} in \sl{Dattātreyayogaśāstra} 140, \emph{Yogabīja} 95 and \emph{Śivasaṃhitā} 3.110 (from which this verse is likely to be derived). Brahmānanda understands \emph{paścima} as referring to the Suṣumṇā (\emph{Jyotsnā} 1.29): \emph{paścimavāhinaṃ paścimena paścimamārgeṇa suṣumṇāmārgeṇa vahatīti paścimavāhī}.
\end{philcomm}

@@

\begin{philcomm}[hp01_030]
%??JM: authorial reworking? 
The source of this verse is unknown, but it has the same elements as the two verses in the \emph{Vasiṣṭhasaṃhitā} (1.76–77), which are themselves derived from earlier Vaiṣṇava sources. The compound \emph{uccāsanaḥ} in the third verse quarter seems to approximate in a somewhat vague way the \emph{Vasiṣṭhasaṃhitā}’s reading \emph{samunnataśiraḥpādaḥ}.

@@

%%%%%%%%%%
\subsection*{1.32}
\begin{translation}[hp01_032]
Lying on one's back on the ground like a corpse is the corpse posture. It removes the fatigue [caused by practising] any \emph{āsana} and relaxes the mind.
\end{translation}

@@

[The yogi] should put the heel at the perineum, firmly place the [other] foot on the penis, hold the face and chest together and the body erect, and [remain] stock-still, his senses restrained, gazing between the brows with his eyes unmoving. This, which breaks open the door to liberation, is called the adept’s pose.
\end{translation}

@@

\begin{testimonia}[hp01_035]
\emph{Yogacintāmaṇi} f. 84v–85r (attr. \emph{Pavanayogasaṅgraha})

@@

Place the left heel on the penis and put the other heel on top: this is \emph{siddhāsana}.\\

@@

\begin{philcomm}[hp01_036]
Svātmārāma's introductory and following remarks to verse 1.36 indicate that he preferred the \emph{siddhāsana} of the \emph{Vivekamārtaṇḍa} over the version taught as \emph{muktāsana} in the \emph{Vasiṣṭhasaṃhitā} and \emph{Yogayājñavalkya}.
\end{philcomm}
%%%%%%%%%%
\subsection*{1.37}
\begin{translation}[hp01_037]
Some call this \emph{siddhāsana}, others know it as \emph{vajrāsana}, a few say it is \emph{muktāsana} and some call it \emph{guptāsana}.
\end{translation}

>>>>

\input{PreambleComm}
%\def\Sone{{\scriptsize{\'S\lower.4ex\hbox{\textrm{\tiny{1}}}}}}
%\def\Soneac{{\scriptsize{\'S\lower.4ex\hbox{\textrm{\tiny{1}}}\kern-2.2pt\lower-1ex\hbox{\textrm{\tiny{ac}}\kern1pt}}}}
%\def\Sonepc{{\scriptsize{\'S\lower.4ex\hbox{\textrm{\tiny{1}}}\kern-2.2pt\lower-1ex\hbox{\textrm{\tiny{pc}}\kern.7pt}}}}
\begin{document}
\pagestyle{HPed}
\begin{ekdosis}
\SetLineation{lineation = none,}

@@

%%%%%%%%%%
\subsection*{1.1}
\begin{translation}[hp01_001]
Homage to the glorious Ādinātha who taught the science of Haṭhayoga which is like a splendid stairway for one who wants to climb to the lofty royal terrace.%Add note re meaning of "science".JM: would much prefer "science" to "system" (vidyā doesn't mean system). I changed "ladder" to "stairway" (a bit more poetic). I'd also prefer "which is like a splendid stairway for one who wants to climb to the lofty royal terrace."
%JB: I much prefer Jim's alternative translation (i.e., "which is like...). But I think we should avoid translating vidyā as `science' because it implies that there was some antecedent to the scientific method in premodern yoga traditions. Also, in light of the current politics around yoga in India, "science of Haṭha" fosters the rather absurd view that ancient Indian knowledge systems anticipated scientific discoveries of the 20th century.  JM: But "system" is wrong. JB: doctrine is better than science.

@@

\begin{philcomm}[hp01_001]

In his commentary {\emph Jyotsnā} on \emph{Haṭhapradīpikā} 1.1 Brahmānanda first states that the
author Svātmārāma starts appropriately with a \emph{maṅgala}, a verse of adoration addressing
Ādinātha, i.e.\ Śiva, but in accordance with his non-sectarian approach leaves room for a second
interpretation of the word as Viṣṇu. A sentence later this is contradicted by his explanantion that Ādinātha, who first taught Yoga, taught it to Pārvatī, which limits the scope to Śiva.

%add comment about yoga replacing mārga and saudha here and in the next two verses.
The reading \emph{°rājasaudham} in the third verse quarter is well attested by the manuscripts, including \alphaOne. However, the most common reading in manuscripts on the lower branches of the stemma is \emph{°rājayogam}, which was accepted by Brahmānanda in \emph{Jyotsnā} 1.1 (see below). It apears that some scribes have made a concerted effort to replace words like \emph{saudha}, \emph{vidyā} and \emph{mārga} with \emph{yoga} in the opening verses of the text (see also 1.1d and 1.3b), even at the expense of the poetic imagery.

In light of the variants, which have lead to our critical edition, Brahmānanda's choices and
interpretation of the second half cannot be upheld. This is Brahmānanda's version and the relevant
portion of his commentary:

\begin{versinnote}
śrīādināthāya namo 'stu tasmai\\ 
\ \ yenopadiṣṭā haṭhayogavidyā |\\
vibhrājate pronnatarājayogam\\ 
\ \ āroḍhum icchor adhirohiṇīva ||1.1||
\end{versinnote}

Our choice of \emph{virājate} in the third verse quarter reflects the relevant manuscripts and rhymes with the following \emph{rāja-}. However, the difference in the sense is so minor that it does alter the meaning.

The main problem in Brahmānanda's interpretation is his choice of \emph{-rājayogaṃ} over
\emph{-rājamsaudhaṃ}, which he explains as follows:

\begin{quote}
  \emph{rājayogaś ca sarvavṛttini\-rodhalakṣaṇo 'samprajñātayogaḥ. tam icchor mumukṣor adhirohiṇīva
  adhi\-ruhyate 'nayety adhirohiṇī niḥśreṇīva vibhrājate viśeṣeṇa bhrājate śobhate. yathā
  pronnatasaudham āroḍhum icchor adhirohiṇy anāyāsena saudha\-prāpi\-kā bhavati evaṃ
  haṭhapradīpikāpi pronnatarājayogam āroḍhum icchor anā\-yāsena rājayogaprāpikā bhavatīti
  upamālaṅkāraḥ.  indravajrākhyaṃ vṛttam.}

Rājayoga is the yoga without cognition,\footnote{This must refer to the \emph{asaṃprajñātasamādhi}
  of the \emph{Yogasūtra} via the equation \emph{yogaḥ samādhiḥ} in the \emph{Bhāṣya} on \emph{Yogasūtra}  1.1.} defined as the stopping of all movements of the mind. To a liberation seeker desiring this, [the science of Haṭhayoga] shines like a ladder.\footnote{The commentary adds etymological explanations: ``\emph{adhirohiṇī} means that by which one ascends, i.e.\ a ladder'', and a synonym for ``shines'', which are not translated here.} Just as a ladder leads someone desiring to ascend to a lofty mansion easily to this mansion, in the same way also the
\emph{haṭhapradīpikā} leads someone desiring the lofty Rājayoga easily to Rājayoga. [In this verse]
the trope is a comparison. The metre is Indravajrā.
\end{quote}

The interpretation makes good sense: Haṭhayoga leads effortlessly to Rājayoga, just as a ladder
leads one to the flat on the top floor. And the comparison as outlined by Brahmānanda has all the
elements deemed necessary by Sanskritic poetology:

\begin{enumerate}
\item Particle expressing a comparison (\emph{upamāvācaka}): \emph{iva} 
\item Standard of comparison (\emph{upamāna}): ladder leading to the palace (\emph{saudhaprāpikā
    adhirohiṇī}) 
\item Thing that is compared (\emph{upameya}): \emph{Haṭhapradīpikā}\footnote{The text itself has
    \emph{haṭhayogavidyā} in that position.} that leads to Rājayoga (\emph{rājayogaprāpikā
    haṭhapradīpikā}) 
\item Common quality  (\emph{samānadharma}): Effortlessness (\emph{anāyāsena})
\end{enumerate}

According to classical poetology a comparison containing all four elements is termed a ``full
comparison'' (\emph{pūrṇopamā}), whereas an elision of one or more elements is called a ``deficient
comparison'' (\emph{luptopamā}). One element that cannot be omitted without losing the comparison
is the \emph{upamāna}. In the verse as given by Brahmānanda this would be ``the ladder''. The
implication is spelt out by Brahmānanda in his commmentary: ``Just as a ladder leads someone
desiring to ascend to a lofty mansion easily leads him to this mansion, in the same way also the
\emph{haṭhapradīpikā} leads someone desiring the lofty Rājayoga easily to Rājayoga.'' However, his
text version spells out only the side of the \emph{upameya}, that is, ``Haṭhayoga leads to
Rājayoga'', but leaves the \emph{upamāna} open to interpretation. His mention of the word
\emph{saudha} in the commentary suggests that this was one of the other options he found in
manuscripts—this reading is very well attested—but was rejected by him. In this way Brahmānanda
makes sure that the text states the obvious, but at the same time it loses part of the comparison,
and it loses its dynamics, which is brought out in the reading \emph{rājasaudha} preferred with
good manuscript evidence in our critical edition. We think that Brahmānanda was eager to state at
the outset the primacy of Rājayoga, and to this end sacrificed the more poetically elegant original
reading that mentions the royal mansion as a sort of metaphor for the ``royal yoga''.

% JM: now saudha is preferred for stemmatic reasons too. I suggest the following shortened form of the note: 
% JB: I prefer the longer note for the digital edition and the shorter one (proposed by Jim) for the printed edition. How many readers will have access to Hanneder 2020 and read the whole article? The longer note nicely summarises a longer discussion in the article.
%The \textalpha\ witnesses read \emph{virājate pronnatarājayogam} while several others have either \emph{vibhrājate} or \emph{°rājasaudham} or both. As explained in Hanneder 2020, \emph{virājate} is stylistically preferable and the comparison between the science of \emph{haṭhayoga} and a stairway works only with \emph{°rājasaudham}.

%The reading \emph{rājasaudha} is preferable for poetical reasons, as explained in Hanneder 2020, p. 128–130. Also, the \emph{Jyotsnā} explains that the \emph{alaṃkāra} here is a comparison (\emph{upamā}), which consists of four elements: (1) a particle expressing the comparison JM: what's happened here? ALso, do we need to repeat what Jürgen has already written elsewhere?
%The Sanskrit poeticians explain that when some elements of a comparison remains unexpressed we get an incomplete (\emph{lupta}) comparison. Often words like \emph{iva} or \emph{yathā} are missing or the common property, but if we read \emph{rājayoga} we lose the \emph{upamāna} (i.e., \emph{rājasaudha}). One part of the comparison should not be missing. But from the perspective of \emph{alaṃkāraśāstra} the verse has a problem that has surely caused the dilemma: it either leaves the \emph{upameya} or the \emph{upamāna} incomplete. Brahmānanda has mentioned both ? \emph{yathā adhirohiṇī saudhaprāpikā bhavati evaṃ haṭhadīpikā rājayogaprāpikā bhavati}, but had to supply \emph{rājasaudha} because his text reads \emph{rājayoga}. But to have the \emph{upameya} in the text is odd. Even Brahmānanda could only know of the image (\emph{rājasaudha}) from the alternative reading he did not accept. However, it seems likely that the author would have included the \emph{upamāna} in the text, as in our critical text, so that the reader would know that the upper terrace of the palace is an image for \emph{rājayoga}. It is further likely that the substitution of \emph{rājayoga} for \emph{rājasaudha} was the result of the tendency to insert the word \emph{yoga} in the opening verses of the text wherever possible, even where it does not fit, as can be seen in 1.2d (\emph{haṭhayogopadiśyate}) and 1.3b (\emph{rājayogam ajānatām}). In the latter case the poetical image has also been lost.

@@

\begin{philcomm}[hp01_003] 
Most witnesses (including \alphaTwo) have \emph{rājayogam ajānatām} (`for those ignorant of Rājayoga') in 1.3b. The reading \emph{rājamārgam ajānatām} (\alphaOne and \alphaThree), “for those unable to find the royal highway”, is more appropriate to the metaphor of being lost in darkness.

%In 1.3d \emph{kṛpākaraḥ} is attested by the most important groups, including \textalpha, of the available manuscripts. The readings \emph{prakāśyate} and \emph{kṣamākaraḥ} are attested by some manuscripts in lower branches of the stemma. Since the context is the author helping yogis who have strayed from the royal path, \emph{kṛpākaraḥ} makes good sense.

As Brahmānanda notes, the compound \emph{kṛpākaraḥ} can be understood as one who is compassionate (\emph{kṛpā} + \emph{kara}) or one who is a mine (i.e., a rich source) of compassion (\emph{kṛpā} + \emph{ākara}). In the Devanagari transmission, the \emph{kṣa} of \emph{kṣamākaraḥ} probably arose as a mistake for \emph{kṛ}.       
\end{philcomm}

%%%%%%%%%%
\subsection*{1.4}
\begin{translation}[hp01_004]
For, Matsyendra, Gorakṣa, and other [perfected yogis] discovered the science of Haṭha, and the yogi Svātmārāma knows it through their favour.
\end{translation}

@@

\begin{philcomm}[hp01_008]  
The name Allamaprabhudeva (sometimes Allama Prabhu Deva or Allama Prabhudeva in secondary literature) is frequently transmitted as \emph{allamaḥ prabhu\-devaḥ} (cf. \alphaTwo), as though it were two names. However, manuscripts \alphaOne, \alphaThree\ and others (e.g., \getsiglum{V3,V8,V13,V16,V22,N24,N26,Jyo}) do not have the \emph{visarga} and write it as one name (i.e., \emph{allamaprabhudevaś ca}). This is also the case in some manuscripts of the \emph{Haṭharatnāvalī} (P,T,t1 in Gharote 2009: 35 n. 8).

@@

The \alphaOne\ and \alphaTwo\ reading of \emph{siddhaḥ kāpālikas} is an exception among the manuscripts and seems too vague to be referring to someone within a lineage. However, Khaṇḍa\-kāpālika is well attested by the manuscripts (including \alphaThree) and this name appears in other literature. Examples include \emph{Kathāsaritsāgara} 121.5 ff. (check??), \emph{Bṛhatkathāmañjarī} 10.45 (check??) and Vajrapāṇi’s \emph{Laghutantraṭīkā}, p.45 (\emph{vīrāḥ khaṇḍakāpālikādayaś caturviṃśatiḥ}). It may be a derogatory name for a Kāpālika, coined perhaps by an outsider and connoting something like a defective Kāpālika in the sense of a ‘part-time’ Kāpālika. Alternatively, it could simply refer to one who uses a broken skull as a bowl.
\end{philcomm}

@@

%The reading of \emph{samāśrayo} in \etaOne\ is metrically faulty. 
The \textalpha\ group omits the second hemistich of this verse, but this was probably the result of eyeskip (i.e., \emph{°maṭho haṭhaḥ} is repeated). Both \emph{°jagatām} and \emph{°yuktānām} are well attested by the collated manuscripts. We have adopted \emph{°jagatāṃ} because it makes good sense with  \emph{ādhārakamaṭhah} in light of the cosmological notion that the tortoise supports all the worlds. This reading may not have been understood by some and was changed in other witnesses to \emph{°yuktānām}, which was adopted by Brahmānanda in \emph{Jyotsnā} 1.10.       
\end{philcomm}

@@

%%%%%%%%%%
\subsection*{1.13}
\begin{translation}[hp01_013]
It has a small door and is without cracks, holes and bumps. It is neither too high nor too low in extent and is thickly smeared with cow dung in the proper way. It is clean, free from all annoyances, pleasing on the outside with a verandah, altar and well, surrounded by a wall: these are the characteristics of the yoga hut as taught by the adept practitioners of Haṭha.%JM remove "it is" etc. which read oddly, thus: With a small door and no cracks, holes and bumps, neither too high nor too low in size, thickly smeared with cow dung in the proper way, clean, free from all annoyances, pleasing on the outside with a verandah, altar and well, surrounded by a wall: these are the characteristics of the yoga hut as taught by the adept practitioners of Haṭha.
\end{translation}

@@

\begin{philcomm}[hp01_013]  
The syntax of this verse is problematic. One would expect the features of the hut, which are listed in the first three quarters of the verse, to be in the nominative case. Then, the words \emph{idaṃ lakṣaṇaṃ} in the fourth quarter would refer back to them. However, the compounds in the first three verse-quarters appear to qualify \emph{lakṣaṇa} as though they were adjectives, and this seems to have been the way the verse was composed.

@@

% (MD: Stemmatically, \emph{piṭaka} ``a basket; a boil" is well attested. Is it perhaps possible to interpret the word as ``a bump"? \emph{a-randhra-garta-piṭaka} would then mean that the floor or walls of the hut are not uneven.) JM: yes, sounds good to me.
%JB Okay I've written the note to reflect this.

@@

%%%%%%%%%%
\subsection*{1.14}
\begin{translation}[hp01_014]
Staying in such a hut, free from all worry, in the way taught by his guru [the yogi] should practise nothing but yoga.
\end{translation}

@@

%%%%%%%%%%
\subsection*{1.19}
\begin{translation}[hp01_019]
Placing the soles of both feet well between the knees and thighs [and] sitting up with the body straight: they call that the auspicious pose (\emph{svastikāsana}).%JM: correctly is tricky. It could mean thoroughly, completely — I've looked at some other usages of samyak and that understanding seems to work well quite often, as it does here. E.g. VS 3.37ab hastābhyāṃ bandhayet samyak karṇādikaraṇāni ca, VM 58cd rasānāṃ śoṣanaṃ samyak mahāmudrābhidhīyate. Or Brahmānanda ad 1.35 susthiram samyak sthiraṃ
% JB thoroughly doesn't make sense here. And is it possible to 'completely' place the feet between the knees and thighs?
\end{translation}

@@

%%%%%%%%%%
\subsection*{1.20}
\begin{translation}[hp01_020]
[The yogi] should place his right heel on the left side of the [lower] back, and the left [heel] on the right [side], in the same way. This is cow-faced pose (\emph{gomukhāsana}), which [looks] like a cow's face.
\end{translation}

@@

%%%%%%%%%%
\subsection*{1.21}
\begin{translation}[hp01_021]
 Fixing one foot on one thigh and placing the [other] thigh on the other foot is called hero pose (\emph{vīrāsana}).
\end{translation}

@@

%%%%%%%%%%
\subsection*{1.22}
\begin{translation}[hp01_022]
Knowers of yoga know that the tortoise pose (\emph{kūrmāsana}) arises by carefully blocking the anus with the ankles crossed.%??Tweak note to reflect change to crossed. 
\end{translation}

@@

%%%%%%%%%%
\subsection*{1.23}
\begin{translation}[hp01_023]
[The yogi] correctly assumes lotus pose, inserts the hands between the knees and thighs, places [the hands] on the ground, and remains in the air. This is the wild cock pose (\emph{kukkuṭāsana}).
\end{translation}

@@

%%%%%%%%%%
\subsection*{1.24}
\begin{translation}[hp01_024]
While maintaining the wild cock pose, [the yogi] binds the neck with the hands and lies [on his back] upturned like a tortoise. This is the upturned tortoise (\emph{uttānakūrmaka}).
\end{translation}

@@

\begin{philcomm}[hp01_024]
The oldest dated manuscript, \etaOne, has \emph{kukkuṭāsanavat kṛtvā}, which is a simpler alternative to the widely attested reading \emph{kukkuṭāsanabandhasthaḥ} (including \alphaTwo\ and \alphaThree), which we have accepted. Since there is no known source for this verse other than the \emph{Haṭhapradīpikā}, it appears that the reading of \etaOne\ was an isolated attempt to simplify the syntax.
\end{philcomm}

%%%%%%%%%%
\subsection*{1.25}
\begin{translation}[hp01_025]
Clasping the big toes with the hands and performing the action of drawing a bow as far as the ear is called the bow pose (\emph{dhanurāsana}).
\end{translation}

@@

%%%%%%%%%%
\subsection*{1.26}
\begin{translation}[hp01_026]
[The yogi] should hold the right foot, which is placed at the base of the left thigh, with the [hand of] the right arm, which is wrapped around the outside of the knee, and remain [like that] with his body twisted. This posture was taught by the revered Matsyendranātha.
\end{translation}

@@

%%%%%%%%%%
\subsection*{1.28}
\begin{translation}[hp01_028]
[The yogi] should stretch out the legs on the ground [as straight] as a stick, hold the toes of both feet with the hands, and practise with the forehead placed on the knees. They call this the back-stretch (\emph{paścimatānam}).
\end{translation}

@@

%%%%%%%%%%
\subsection*{1.29}
\begin{translation}[hp01_029]
Foremost among \emph{āsana}s, the back-stretch thus makes the breath flow to the rear (i.e.~in the central channel), increases the digestive fire, makes the belly thin and prevents diseases in men.
\end{translation}

@@

\begin{philcomm}[hp01_029]
 The use of the word \emph{paścima} to mean the central channel is found at \emph{Yogabīja} 95 (\emph{paścimamārgataḥ}), 108 (\emph{paścime pathi}), 117 (\emph{paścimadvāramārgeṇa}) and 121 (\emph{paścimaṃ}). Cf.~the usages of \emph{paścima\-mārga} in \sl{Dattātreyayogaśāstra} 140 and \emph{Śivasaṃhitā} 3.110 (from which this verse is likely to be derived). Brahmānanda understands \emph{paścima} as referring to the Suṣumṇā (\emph{Jyotsnā} 1.29): \emph{paścimavāhinaṃ paścimena paścimamārgeṇa suṣumṇāmārgeṇa vahatīti paścimavāhī}.
\end{philcomm}

@@

\begin{philcomm}[hp01_030]
%
The source of this verse is unknown, but it has the same elements as the two verses in the \emph{Vasiṣṭhasaṃhitā} (1.76–77), which are themselves derived from earlier Vaiṣṇava sources. The compound \emph{uccāsanaḥ} in the third verse quarter seems to approximate in a somewhat vague way the \emph{Vasiṣṭhasaṃhitā}’s reading \emph{samunnataśiraḥpādaḥ}.

@@

%%%%%%%%%%
\subsection*{1.32}
\begin{translation}[hp01_032]
Lying with one's back on the ground like a corpse is the corpse posture. It removes the fatigue [caused by practising] any \emph{āsana} and calms the mind.
\end{translation}

@@

[The yogi] should put the heel at the perineum, firmly place the [other] foot on the penis, focus the mind, hold the body erect and [remain] motionless, his senses restrained, gazing between the brows with his eyes unmoving. This, which breaks open the door to liberation, is called the adept’s pose.
\end{translation}

@@

\begin{testimonia}[hp01_035]
\emph{Yogacintāmaṇi} f.~84v–85r (attr.~\emph{Pavanayogasaṅgraha})

@@

Place the left heel on the penis and put the other heel on top: this is the adept's pose (\emph{siddhāsana}).\\

@@

\begin{philcomm}[hp01_036]
Svātmārāma's introductory and following remarks to verse 1.36 indicate that he preferred the \emph{siddhāsana} of the \emph{Vivekamārtaṇḍa} over the version taught as \emph{muktāsana} in the \emph{Vasiṣṭhasaṃhitā} and \emph{Yogayājñavalkya}.
\end{philcomm}
%%%%%%%%%%
\subsection*{1.37}
\begin{translation}[hp01_037]
Some call this the adept's pose (\emph{siddhāsana}), others know it as the thunderbolt pose (\emph{vajrāsana}), a few say it is the pose of the liberated (\emph{muktāsana}) and some call it the secret pose (\emph{guptāsana}).
\end{translation}
